---
title: "Nanopore Full-Length 16S: Emu Figures"
author: "Kalan Lab"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

This notebook is intentionally divided into small chunks so that every data
transformation and graph can be inspected in RStudio. It uses only samples
listed in `metadata/sample_name_key.tsv`. Sample 16 is listed there, but its
reviewed exploratory outputs are integrated separately because it was excluded
from the original cross-sectional batch run.

# 1. Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 150
)

required_packages <- c(
  "readr", "dplyr", "tidyr", "ggplot2", "stringr", "scales", "knitr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install these R packages before continuing: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(scales)
library(knitr)

project_root <- normalizePath(".", mustWork = TRUE)
path <- function(...) file.path(project_root, ...)

figure_dir <- path("results", "figures_rmarkdown")
table_dir <- path("results", "combined_rmarkdown")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
```

# 2. Fixed species order and colors

```{r species-settings}
species_order <- c(
  "Bacteroides thetaiotaomicron",
  "Cutibacterium acnes",
  "Fusobacterium nucleatum",
  "Prevotella melaninogenica",
  "Veillonella parvula"
)

unknown_label <- "Unknown / other"
stack_order <- c(species_order, unknown_label)

species_colors <- c(
  "Bacteroides thetaiotaomicron" = "#0072B2",
  "Cutibacterium acnes" = "#CC79A7",
  "Fusobacterium nucleatum" = "#D55E00",
  "Prevotella melaninogenica" = "#009E73",
  "Veillonella parvula" = "#E69F00",
  "Unknown / other" = "#8A8A8A"
)

assignment_method_colors <- c(
  "Custom Emu: five references" = "#0072B2",
  "Broad Emu: any classified taxon" = "#D55E00"
)

minimap_status_colors <- c(
  "Confident" = "#009E73",
  "Ambiguous" = "#E69F00",
  "Unassigned" = "#7A7A7A"
)

theme_16s <- theme_bw(base_size = 11) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1),
    legend.position = "bottom",
    legend.text = element_text(size = 9)
  )
```

# 3. Load the sample-name key

Edit `metadata/sample_name_key.tsv` to change a display name or the plotting
order. Samples absent from this file are excluded from all graphs below.

```{r load-sample-key}
sample_key_file <- path("metadata", "sample_name_key.tsv")
if (!file.exists(sample_key_file)) {
  stop("Missing sample-name key: ", sample_key_file)
}

sample_key <- read_tsv(sample_key_file, show_col_types = FALSE) %>%
  arrange(display_order)

if (anyDuplicated(sample_key$sample_id)) {
  stop("Duplicate sample_id values in metadata/sample_name_key.tsv")
}
if (anyDuplicated(sample_key$display_name)) {
  stop("Duplicate display_name values in metadata/sample_name_key.tsv")
}
if (anyDuplicated(sample_key$display_order)) {
  stop("Duplicate display_order values in metadata/sample_name_key.tsv")
}

selected_sample_ids <- sample_key$sample_id
selected_sample_names <- sample_key$display_name

kable(sample_key, caption = "Samples retained for the R analysis")
```

# 4. Load Emu and expected-ratio tables

```{r load-data}
custom_file <- path(
  "results", "combined",
  "custom_emu_expected_species_abundance_renormalized.tsv"
)
broad_species_file <- path(
  "results", "combined", "broad_emu_species_abundance.tsv"
)
broad_accounting_file <- path(
  "results", "combined", "broad_emu_read_accounting.tsv"
)
minimap_status_file <- path(
  "results", "combined", "minimap2_read_status_summary.tsv"
)
minimap_confident_file <- path(
  "results", "combined", "minimap2_confident_reference_abundance.tsv"
)
expected_file <- path("metadata", "expected_ratios_template.tsv")
sample16_id <- "2PJGZM_16_sample_16"
sample16_custom_file <- path(
  "results",
  "exploratory_sample16_standard_filters",
  "emu",
  "sample16_exploratory_rel-abundance.tsv"
)
sample16_broad_file <- path(
  "results",
  "exploratory_sample16_standard_filters",
  "broad_emu_retry1",
  "sample16_broad_exploratory_rel-abundance.tsv"
)
sample16_minimap_file <- path(
  "results",
  "minimap2",
  sample16_id,
  paste0(sample16_id, ".per_read_assignments.tsv")
)

required_files <- c(
  custom_file,
  broad_species_file,
  broad_accounting_file,
  minimap_status_file,
  minimap_confident_file,
  expected_file,
  sample16_custom_file,
  sample16_broad_file,
  sample16_minimap_file
)
missing_files <- required_files[!file.exists(required_files)]

if (length(missing_files) > 0) {
  stop("Missing input file(s):\n", paste(missing_files, collapse = "\n"))
}

custom_raw <- read_tsv(custom_file, show_col_types = FALSE)
broad_species_raw <- read_tsv(broad_species_file, show_col_types = FALSE)
broad_accounting_raw <- read_tsv(
  broad_accounting_file,
  show_col_types = FALSE
)
minimap_status_raw <- read_tsv(
  minimap_status_file,
  show_col_types = FALSE
)
minimap_confident_raw <- read_tsv(
  minimap_confident_file,
  show_col_types = FALSE
)
expected_raw <- read_tsv(expected_file, show_col_types = FALSE)
```

## Integrate the separately processed sample 16

Sample 16 was processed separately after review of its atypical raw
read-length distribution. This chunk converts its native Emu outputs and
minimap2 per-read assignments into the same structures as the combined
primary-analysis tables. Existing sample 16 rows are replaced, making this
chunk safe to rerun after future combined tables are regenerated.

```{r integrate-sample16-results}
special_emu_rows <- c(
  "unmapped", "mapped_filtered", "mapped_unclassified"
)

sample16_custom_native <- read_tsv(
  sample16_custom_file,
  show_col_types = FALSE,
  na = c("", "NA")
)
sample16_custom_total_reads <- sum(
  sample16_custom_native$`estimated counts`,
  na.rm = TRUE
)
sample16_custom_expected_reads <- sample16_custom_native %>%
  filter(species %in% species_order) %>%
  summarise(total = sum(`estimated counts`, na.rm = TRUE)) %>%
  pull(total)

sample16_custom_rows <- sample16_custom_native %>%
  filter(species %in% species_order) %>%
  transmute(
    sample_id = sample16_id,
    species,
    emu_abundance_pct_renormalized_among_expected =
      100 * `estimated counts` / sample16_custom_expected_reads,
    emu_abundance_pct_all_usable_reads =
      100 * `estimated counts` / sample16_custom_total_reads,
    estimated_counts = `estimated counts`,
    source_file = sample16_custom_file
  ) %>%
  complete(
    sample_id = sample16_id,
    species = species_order,
    fill = list(
      emu_abundance_pct_renormalized_among_expected = 0,
      emu_abundance_pct_all_usable_reads = 0,
      estimated_counts = 0,
      source_file = sample16_custom_file
    )
  )

custom_raw <- custom_raw %>%
  filter(sample_id != sample16_id) %>%
  bind_rows(sample16_custom_rows)

sample16_broad_native <- read_tsv(
  sample16_broad_file,
  show_col_types = FALSE,
  na = c("", "NA")
)

sample16_broad_taxa <- sample16_broad_native %>%
  filter(
    !tax_id %in% special_emu_rows,
    !is.na(species),
    species != ""
  )

sample16_broad_total_reads <- sum(
  sample16_broad_native$`estimated counts`,
  na.rm = TRUE
)
sample16_broad_classified_reads <- sum(
  sample16_broad_taxa$`estimated counts`,
  na.rm = TRUE
)
sample16_expected_reads <- sample16_broad_taxa %>%
  filter(species %in% species_order) %>%
  summarise(total = sum(`estimated counts`, na.rm = TRUE)) %>%
  pull(total)
sample16_unexpected_reads <- sample16_broad_taxa %>%
  filter(!species %in% species_order) %>%
  summarise(total = sum(`estimated counts`, na.rm = TRUE)) %>%
  pull(total)

sample16_special_count <- function(category) {
  sample16_broad_native %>%
    filter(tax_id == category) %>%
    summarise(total = sum(`estimated counts`, na.rm = TRUE)) %>%
    pull(total)
}

sample16_unmapped_reads <- sample16_special_count("unmapped")
sample16_filtered_reads <- sample16_special_count("mapped_filtered")
sample16_mapped_unclassified_reads <-
  sample16_special_count("mapped_unclassified")
sample16_taxonomy_unclassified_reads <- sample16_broad_native %>%
  filter(
    !tax_id %in% special_emu_rows,
    is.na(species) | species == ""
  ) %>%
  summarise(total = sum(`estimated counts`, na.rm = TRUE)) %>%
  pull(total)

sample16_broad_rows <- sample16_broad_taxa %>%
  transmute(
    sample_id = sample16_id,
    tax_id = as.numeric(tax_id),
    species,
    genus,
    taxon_category = if_else(
      species %in% species_order,
      "expected_species",
      "unexpected_taxon"
    ),
    estimated_counts = `estimated counts`,
    emu_abundance_pct_among_classified =
      100 * `estimated counts` / sample16_broad_classified_reads,
    abundance_pct_all_usable_reads =
      100 * `estimated counts` / sample16_broad_total_reads,
    source_file = sample16_broad_file
  )

broad_species_raw <- broad_species_raw %>%
  filter(sample_id != sample16_id) %>%
  bind_rows(sample16_broad_rows)

sample16_broad_accounting <- tibble(
  sample_id = sample16_id,
  usable_reads = sample16_broad_total_reads,
  emu_classified_estimated_reads = sample16_broad_classified_reads,
  expected_species_estimated_reads = sample16_expected_reads,
  unexpected_species_estimated_reads = sample16_unexpected_reads,
  mapped_unclassified_reads = sample16_mapped_unclassified_reads,
  mapped_filtered_reads = sample16_filtered_reads,
  unmapped_reads = sample16_unmapped_reads,
  taxonomy_unclassified_reads = sample16_taxonomy_unclassified_reads,
  total_non_expected_pct_all_usable_reads = 100 * (
    sample16_unexpected_reads +
      sample16_mapped_unclassified_reads +
      sample16_filtered_reads +
      sample16_unmapped_reads +
      sample16_taxonomy_unclassified_reads
  ) / sample16_broad_total_reads,
  status = "completed_exploratory"
)

broad_accounting_raw <- broad_accounting_raw %>%
  filter(sample_id != sample16_id) %>%
  bind_rows(sample16_broad_accounting)

sample16_minimap_native <- read_tsv(
  sample16_minimap_file,
  show_col_types = FALSE,
  na = c("", "NA")
)
sample16_minimap_total_reads <- nrow(sample16_minimap_native)
sample16_minimap_confident_reads <- sum(
  sample16_minimap_native$assignment_status == "confident"
)
sample16_minimap_ambiguous_reads <- sum(
  sample16_minimap_native$assignment_status == "ambiguous"
)
sample16_minimap_unassigned_reads <- sum(
  sample16_minimap_native$assignment_status == "unassigned"
)

sample16_minimap_status <- tibble(
  sample_id = sample16_id,
  total_reads = sample16_minimap_total_reads,
  confident_reads = sample16_minimap_confident_reads,
  ambiguous_reads = sample16_minimap_ambiguous_reads,
  unassigned_reads = sample16_minimap_unassigned_reads,
  confident_pct =
    100 * confident_reads / total_reads,
  ambiguous_pct =
    100 * ambiguous_reads / total_reads,
  unassigned_pct =
    100 * unassigned_reads / total_reads
)

minimap_status_raw <- minimap_status_raw %>%
  filter(sample_id != sample16_id) %>%
  bind_rows(sample16_minimap_status)

sample16_minimap_confident <- sample16_minimap_native %>%
  filter(assignment_status == "confident") %>%
  count(best_reference, name = "confident_reads") %>%
  transmute(
    sample_id = sample16_id,
    reference = best_reference,
    confident_reads,
    percentage_all_input_reads =
      100 * confident_reads / sample16_minimap_total_reads,
    percentage_among_confident_reads =
      100 * confident_reads / sample16_minimap_confident_reads
  )

minimap_confident_raw <- minimap_confident_raw %>%
  filter(sample_id != sample16_id) %>%
  bind_rows(sample16_minimap_confident)

tibble(
  method = c("Custom Emu", "Broad Emu", "Minimap2"),
  sample16_input_reads = c(
    sample16_custom_total_reads,
    sample16_broad_total_reads,
    sample16_minimap_total_reads
  ),
  assigned_or_confident_reads = c(
    sample16_custom_expected_reads,
    sample16_broad_classified_reads,
    sample16_minimap_confident_reads
  )
) %>%
  kable(
    digits = 3,
    caption = "Sample 16 supplemental-result integration check"
  )
```

## Inspect the raw columns

```{r inspect-inputs}
list(
  custom_columns = names(custom_raw),
  broad_species_columns = names(broad_species_raw),
  broad_accounting_columns = names(broad_accounting_raw),
  minimap_status_columns = names(minimap_status_raw),
  minimap_confident_columns = names(minimap_confident_raw),
  expected_ratio_columns = names(expected_raw)
)
```

```{r validate-selected-samples}
missing_custom <- setdiff(selected_sample_ids, unique(custom_raw$sample_id))
missing_broad <- setdiff(
  selected_sample_ids,
  unique(broad_accounting_raw$sample_id)
)
missing_minimap <- setdiff(
  selected_sample_ids,
  unique(minimap_status_raw$sample_id)
)

if (length(missing_custom) > 0) {
  stop(
    "Selected samples missing from custom Emu results: ",
    paste(missing_custom, collapse = ", ")
  )
}
if (length(missing_broad) > 0) {
  stop(
    "Selected samples missing from broad Emu results: ",
    paste(missing_broad, collapse = ", ")
  )
}
if (length(missing_minimap) > 0) {
  stop(
    "Selected samples missing from minimap2 results: ",
    paste(missing_minimap, collapse = ", ")
  )
}

tibble(
  data_source = c(
    "Custom Emu",
    "Broad Emu accounting",
    "Minimap2 read status"
  ),
  selected_samples_found = c(
    sum(selected_sample_ids %in% unique(custom_raw$sample_id)),
    sum(selected_sample_ids %in% unique(broad_accounting_raw$sample_id)),
    sum(selected_sample_ids %in% unique(minimap_status_raw$sample_id))
  ),
  selected_samples_expected = length(selected_sample_ids)
) %>%
  kable(caption = "Input completeness check")
```

# 5. Custom five-reference Emu stacked abundance

This graph uses the custom database and renormalizes the five expected species
to 100%. It describes consortium composition among reads assigned to those five
references. It does not display unassigned reads.

## Prepare custom Emu plotting data

```{r prepare-custom-data}
custom_plot_data <- custom_raw %>%
  filter(sample_id %in% selected_sample_ids) %>%
  select(
    sample_id,
    species,
    abundance_pct = emu_abundance_pct_renormalized_among_expected
  ) %>%
  complete(
    sample_id = selected_sample_ids,
    species = species_order,
    fill = list(abundance_pct = 0)
  ) %>%
  left_join(sample_key, by = "sample_id") %>%
  mutate(
    display_name = factor(display_name, levels = selected_sample_names),
    species = factor(species, levels = species_order)
  ) %>%
  arrange(display_order, species)

custom_sum_check <- custom_plot_data %>%
  group_by(sample_id, display_name) %>%
  summarise(total_pct = sum(abundance_pct), .groups = "drop")

kable(
  custom_sum_check,
  digits = 3,
  caption = "Custom Emu percentage-sum check"
)
```

## Draw the custom Emu graph

```{r plot-custom-stacked, fig.width=12, fig.height=6}
custom_stacked_plot <- ggplot(
  custom_plot_data,
  aes(x = display_name, y = abundance_pct, fill = species)
) +
  geom_col(width = 0.82, color = "white", linewidth = 0.15) +
  scale_fill_manual(values = species_colors[species_order], drop = FALSE) +
  scale_y_continuous(
    breaks = seq(0, 100, by = 20),
    labels = label_number(suffix = "%"),
    expand = expansion(mult = c(0, 0))
  ) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(
    x = NULL,
    y = "Abundance among five expected species",
    fill = NULL,
    title = "Custom five-reference Emu"
  ) +
  theme_16s

custom_stacked_plot
```

```{r save-custom-stacked}
write_tsv(
  custom_plot_data %>%
    mutate(
      display_name = as.character(display_name),
      species = as.character(species)
    ),
  file.path(table_dir, "custom_emu_stacked_plot_data.tsv")
)

ggsave(
  file.path(figure_dir, "custom_emu_stacked_selected_samples.png"),
  custom_stacked_plot,
  width = 12,
  height = 6,
  dpi = 300
)
ggsave(
  file.path(figure_dir, "custom_emu_stacked_selected_samples.pdf"),
  custom_stacked_plot,
  width = 12,
  height = 6
)
```

## P0 versus P1 custom Emu composition

This focused comparison relabels sample 1 as `P0` and sample 16 as `P1`.
Percentages are renormalized among reads assigned to the five expected species,
as in the main custom-Emu graph. P0 had substantially lower read support and
was excluded from the current main figures, so its composition should be
interpreted cautiously.

```{r prepare-p0-p1-custom-emu}
p0_p1_key <- tribble(
  ~sample_id, ~passage, ~passage_order,
  "2PJGZM_1_sample_1", "P0", 1,
  sample16_id, "P1", 2
)

missing_p0_p1 <- setdiff(p0_p1_key$sample_id, unique(custom_raw$sample_id))
if (length(missing_p0_p1) > 0) {
  stop(
    "P0/P1 samples missing from custom Emu results: ",
    paste(missing_p0_p1, collapse = ", ")
  )
}

p0_p1_custom_data <- custom_raw %>%
  filter(sample_id %in% p0_p1_key$sample_id) %>%
  select(
    sample_id,
    species,
    abundance_pct = emu_abundance_pct_renormalized_among_expected
  ) %>%
  complete(
    sample_id = p0_p1_key$sample_id,
    species = species_order,
    fill = list(abundance_pct = 0)
  ) %>%
  left_join(p0_p1_key, by = "sample_id") %>%
  mutate(
    passage = factor(passage, levels = c("P0", "P1")),
    species = factor(species, levels = species_order)
  ) %>%
  arrange(passage_order, species)

kable(
  p0_p1_custom_data,
  digits = 3,
  caption = "Custom Emu composition used for the P0-P1 comparison"
)
```

```{r plot-p0-p1-custom-emu, fig.width=9, fig.height=5.5}
p0_p1_custom_plot <- ggplot(
  p0_p1_custom_data,
  aes(x = passage, y = abundance_pct, fill = species)
) +
  geom_col(width = 0.68, color = "white", linewidth = 0.2) +
  scale_fill_manual(values = species_colors[species_order], drop = FALSE) +
  scale_y_continuous(
    breaks = seq(0, 100, by = 20),
    labels = label_number(suffix = "%"),
    expand = expansion(mult = c(0, 0))
  ) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(
    x = NULL,
    y = "Abundance among five expected species",
    fill = NULL,
    title = "Custom Emu composition: P0 versus P1",
    subtitle = "P0 = sample 1; P1 = sample 16",
    caption = "P0 has lower sequencing depth and should be interpreted cautiously."
  ) +
  theme_16s +
  guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.text = element_text(size = 8)
  )

p0_p1_custom_plot
```

```{r save-p0-p1-custom-emu}
write_tsv(
  p0_p1_custom_data %>%
    mutate(
      passage = as.character(passage),
      species = as.character(species)
    ),
  file.path(table_dir, "p0_p1_custom_emu_stacked.tsv")
)

ggsave(
  file.path(figure_dir, "p0_p1_custom_emu_stacked.png"),
  p0_p1_custom_plot,
  width = 9,
  height = 5.5,
  dpi = 300
)
```

## Sample 1 organism percentages

This single-sample graph shows the same targeted custom-Emu composition for
sample 1 (`P0`) as separate organism bars. Labels are retained for zero and
low-percentage organisms that are difficult to read in a stacked bar.

```{r prepare-sample1-custom-emu}
sample1_custom_data <- p0_p1_custom_data %>%
  filter(as.character(passage) == "P0") %>%
  transmute(
    sample_id,
    passage = as.character(passage),
    species = as.character(species),
    abundance_pct,
    species_plot = factor(species, levels = rev(species_order))
  )

kable(
  sample1_custom_data %>%
    select(sample_id, passage, species, abundance_pct),
  digits = 3,
  caption = "Custom Emu organism percentages for sample 1 (P0)"
)
```

```{r plot-sample1-custom-emu, fig.width=8, fig.height=5.5}
sample1_custom_plot <- ggplot(
  sample1_custom_data,
  aes(x = abundance_pct, y = species_plot, fill = species)
) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(
    aes(label = sprintf("%.2f%%", abundance_pct)),
    hjust = -0.15,
    size = 3.8
  ) +
  scale_fill_manual(values = species_colors[species_order], drop = FALSE) +
  scale_x_continuous(
    breaks = seq(0, 100, by = 20),
    labels = label_number(suffix = "%"),
    limits = c(0, 100),
    expand = expansion(mult = c(0, 0.06))
  ) +
  labs(
    x = "Abundance among five expected species",
    y = NULL,
    title = "Sample 1 (P0): custom Emu composition",
    subtitle = "Relative abundance among reads assigned to the five custom references",
    caption = "Sample 1 has low sequencing depth; small percentages should be interpreted cautiously."
  ) +
  theme_16s +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.text.y = element_text(face = "italic")
  )

sample1_custom_plot
```

```{r save-sample1-custom-emu}
write_tsv(
  sample1_custom_data %>%
    select(sample_id, passage, species, abundance_pct),
  file.path(table_dir, "sample1_p0_custom_emu_percentages.tsv")
)

ggsave(
  file.path(figure_dir, "sample1_p0_custom_emu_percentages.png"),
  sample1_custom_plot,
  width = 8,
  height = 5.5,
  dpi = 300
)
```

# 6. Broad Emu: five known species plus unknown/other

This graph uses assignments from the published broad Emu database, not the
custom FASTA database. The five named segments are exact species assignments
from broad Emu. `Unknown / other` is everything not assigned to one of those
five exact species, including unexpected taxa, unmapped reads, filtered
alignments, and unclassified reads.

No broad Emu rerun is needed when
`results/combined/broad_emu_species_abundance.tsv` and
`results/combined/broad_emu_read_accounting.tsv` contain every selected sample.

## Prepare broad Emu known-species percentages

```{r prepare-broad-known}
broad_known <- broad_species_raw %>%
  filter(
    sample_id %in% selected_sample_ids,
    species %in% species_order
  ) %>%
  group_by(sample_id, species) %>%
  summarise(
    abundance_pct = sum(abundance_pct_all_usable_reads),
    .groups = "drop"
  ) %>%
  complete(
    sample_id = selected_sample_ids,
    species = species_order,
    fill = list(abundance_pct = 0)
  )

kable(
  broad_known %>%
    filter(sample_id %in% head(selected_sample_ids, 4)),
  digits = 3,
  caption = "Example broad Emu exact-species percentages"
)
```

## Calculate unknown/other

```{r prepare-broad-unknown}
broad_unknown <- broad_accounting_raw %>%
  filter(sample_id %in% selected_sample_ids) %>%
  transmute(
    sample_id,
    species = unknown_label,
    abundance_pct = total_non_expected_pct_all_usable_reads
  )

broad_plot_data <- bind_rows(broad_known, broad_unknown) %>%
  left_join(sample_key, by = "sample_id") %>%
  mutate(
    display_name = factor(display_name, levels = selected_sample_names),
    species = factor(species, levels = stack_order)
  ) %>%
  arrange(display_order, species)

broad_sum_check <- broad_plot_data %>%
  group_by(sample_id, display_name) %>%
  summarise(total_pct = sum(abundance_pct), .groups = "drop")

kable(
  broad_sum_check,
  digits = 3,
  caption = "Broad Emu percentage-sum check"
)
```

## Inspect what contributes to unknown/other

```{r inspect-broad-unknown-components}
broad_unknown_components <- broad_accounting_raw %>%
  filter(sample_id %in% selected_sample_ids) %>%
  left_join(sample_key, by = "sample_id") %>%
  transmute(
    display_order,
    display_name,
    usable_reads,
    unexpected_taxa_pct =
      100 * unexpected_species_estimated_reads / usable_reads,
    mapped_unclassified_pct =
      100 * mapped_unclassified_reads / usable_reads,
    alignment_filtered_pct =
      100 * mapped_filtered_reads / usable_reads,
    unmapped_pct =
      100 * unmapped_reads / usable_reads,
    taxonomy_unclassified_pct =
      100 * taxonomy_unclassified_reads / usable_reads,
    total_unknown_or_other_pct =
      total_non_expected_pct_all_usable_reads
  ) %>%
  arrange(display_order)

kable(
  broad_unknown_components %>% select(-display_order),
  digits = 3,
  caption = "Components of broad Emu unknown/other"
)
```

## Draw the broad Emu graph

```{r plot-broad-stacked, fig.width=12, fig.height=6}
broad_stacked_plot <- ggplot(
  broad_plot_data,
  aes(x = display_name, y = abundance_pct, fill = species)
) +
  geom_col(width = 0.82, color = "white", linewidth = 0.15) +
  scale_fill_manual(values = species_colors[stack_order], drop = FALSE) +
  scale_y_continuous(
    breaks = seq(0, 100, by = 20),
    labels = label_number(suffix = "%"),
    expand = expansion(mult = c(0, 0))
  ) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(
    x = NULL,
    y = "Abundance among all QC-passing reads",
    fill = NULL,
    title = "Published broad Emu database",
    subtitle = "Five exact species assignments plus unknown/other"
  ) +
  theme_16s

broad_stacked_plot
```

```{r save-broad-stacked}
write_tsv(
  broad_plot_data %>%
    mutate(
      display_name = as.character(display_name),
      species = as.character(species)
    ),
  file.path(table_dir, "broad_emu_known_plus_unknown_plot_data.tsv")
)
write_tsv(
  broad_unknown_components,
  file.path(table_dir, "broad_emu_unknown_components.tsv")
)

ggsave(
  file.path(
    figure_dir,
    "broad_emu_five_species_plus_unknown_selected_samples.png"
  ),
  broad_stacked_plot,
  width = 12,
  height = 6,
  dpi = 300
)
ggsave(
  file.path(
    figure_dir,
    "broad_emu_five_species_plus_unknown_selected_samples.pdf"
  ),
  broad_stacked_plot,
  width = 12,
  height = 6
)
```

# 7. Samples 4-7: custom Emu versus expected CFU ratios

The expected percentages are CFU-adjusted. They were calculated from the input
volume ratios and the measured strain-specific log10 CFU/mL values. The
observed custom Emu values are renormalized among the five expected species, so
both bars sum to 100%.

## Convert expected percentages to long format

```{r prepare-expected-data}
comparison_ids <- paste0("2PJGZM_", 4:7, "_sample_", 4:7)

expected_long <- expected_raw %>%
  filter(sample_id %in% comparison_ids) %>%
  select(sample_id, ratio_basis, ends_with("_expected_pct")) %>%
  pivot_longer(
    cols = ends_with("_expected_pct"),
    names_to = "species",
    values_to = "abundance_pct"
  ) %>%
  mutate(
    species = str_remove(species, "_expected_pct$"),
    species = str_replace_all(species, "_", " ")
  )

if (!all(expected_long$ratio_basis == "CFU")) {
  stop("Samples 4-7 are expected to use CFU-adjusted expected percentages")
}

expected_long %>%
  group_by(sample_id) %>%
  summarise(expected_total_pct = sum(abundance_pct), .groups = "drop") %>%
  kable(digits = 3, caption = "Expected-ratio percentage-sum check")
```

## Combine expected and observed values

```{r combine-expected-observed}
observed_4_7 <- custom_plot_data %>%
  filter(sample_id %in% comparison_ids) %>%
  transmute(
    sample_id,
    species = as.character(species),
    abundance_pct,
    result_type = "Observed: custom Emu"
  )

expected_4_7 <- expected_long %>%
  transmute(
    sample_id,
    species,
    abundance_pct,
    result_type = "Expected: CFU-adjusted"
  )

comparison_plot_data <- bind_rows(expected_4_7, observed_4_7) %>%
  left_join(sample_key, by = "sample_id") %>%
  mutate(
    display_name = factor(
      display_name,
      levels = sample_key %>%
        filter(sample_id %in% comparison_ids) %>%
        arrange(display_order) %>%
        pull(display_name)
    ),
    species = factor(species, levels = species_order),
    result_type = factor(
      result_type,
      levels = c("Expected: CFU-adjusted", "Observed: custom Emu")
    )
  )

comparison_table <- comparison_plot_data %>%
  mutate(
    display_name = as.character(display_name),
    species = as.character(species),
    result_type = as.character(result_type)
  ) %>%
  select(sample_id, display_name, species, result_type, abundance_pct) %>%
  pivot_wider(
    names_from = result_type,
    values_from = abundance_pct
  ) %>%
  mutate(
    difference_pct_points =
      `Observed: custom Emu` - `Expected: CFU-adjusted`
  )

kable(
  comparison_table,
  digits = 2,
  caption = "Expected and observed percentages for samples 4-7"
)
```

## Draw paired stacked bars

```{r plot-expected-observed, fig.width=11, fig.height=6}
expected_observed_plot <- ggplot(
  comparison_plot_data,
  aes(x = result_type, y = abundance_pct, fill = species)
) +
  geom_col(width = 0.72, color = "white", linewidth = 0.15) +
  facet_wrap(
    vars(display_name),
    nrow = 1,
    labeller = label_wrap_gen(width = 18)
  ) +
  scale_fill_manual(values = species_colors[species_order], drop = FALSE) +
  scale_y_continuous(
    breaks = seq(0, 100, by = 20),
    labels = label_number(suffix = "%"),
    expand = expansion(mult = c(0, 0))
  ) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(
    x = NULL,
    y = "Composition",
    fill = NULL,
    title = "Expected CFU composition versus custom Emu",
    subtitle = "Samples 4-7; observed values renormalized among five species"
  ) +
  theme_16s +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1),
    strip.text = element_text(face = "bold")
  )

expected_observed_plot
```

```{r save-expected-observed}
write_tsv(
  comparison_table,
  file.path(table_dir, "samples_4_7_expected_vs_custom_emu.tsv")
)

ggsave(
  file.path(
    figure_dir,
    "samples_4_7_expected_cfu_vs_custom_emu.png"
  ),
  expected_observed_plot,
  width = 11,
  height = 6,
  dpi = 300
)
ggsave(
  file.path(
    figure_dir,
    "samples_4_7_expected_cfu_vs_custom_emu.pdf"
  ),
  expected_observed_plot,
  width = 11,
  height = 6
)
```

# 8. Known-mixture validation metrics

For each known mixture, the error is calculated from the five expected and
observed species percentages:

* **MAE across all five species** is the mean absolute percentage-point
  difference. Lower values indicate better agreement.
* **MAE among expected-present species** excludes organisms with an expected
  percentage of zero. This prevents correctly absent organisms from making a
  sparse mixture appear more accurate than its recovered organisms actually
  are.
* **RMSE** gives more weight to large organism-specific errors.
* **Bray-Curtis similarity** ranges from 0 to 1, with 1 indicating identical
  expected and observed compositions.

When both compositions sum to 100% and the same five species are used,
Bray-Curtis similarity is calculated as:

`1 - sum(abs(observed - expected)) / 200`

MAE and Bray-Curtis similarity therefore summarize the same absolute
differences on different scales; they should not be treated as independent
validation evidence.

## Calculate per-mixture validation statistics

```{r calculate-known-mixture-validation}
known_mixture_validation <- comparison_table %>%
  left_join(
    sample_key %>% select(sample_id, display_order),
    by = "sample_id"
  ) %>%
  group_by(sample_id, display_name, display_order) %>%
  summarise(
    expected_present_species =
      sum(`Expected: CFU-adjusted` > 0),
    mae_all_five_pct_points =
      mean(abs(difference_pct_points)),
    mae_expected_present_pct_points =
      mean(
        abs(
          difference_pct_points[
            `Expected: CFU-adjusted` > 0
          ]
        )
      ),
    rmse_pct_points =
      sqrt(mean(difference_pct_points^2)),
    bray_curtis_dissimilarity =
      sum(abs(difference_pct_points)) /
      sum(
        `Expected: CFU-adjusted` +
        `Observed: custom Emu`
      ),
    bray_curtis_similarity =
      1 - bray_curtis_dissimilarity,
    maximum_absolute_error_pct_points =
      max(abs(difference_pct_points)),
    .groups = "drop"
  ) %>%
  mutate(
    validation_status = if_else(
      mae_all_five_pct_points <= 10 &
        bray_curtis_similarity >= 0.80,
      "Pass proposed semi-quantitative criteria",
      "Review"
    )
  ) %>%
  arrange(display_order)

kable(
  known_mixture_validation %>%
    select(
      display_name,
      expected_present_species,
      mae_all_five_pct_points,
      mae_expected_present_pct_points,
      rmse_pct_points,
      bray_curtis_similarity,
      maximum_absolute_error_pct_points,
      validation_status
    ),
  digits = 3,
  caption = "Known-mixture validation statistics"
)
```

## Display overall known-mixture statistics

```{r summarize-known-mixture-validation}
known_mixture_overall_stats <- comparison_table %>%
  summarise(
    samples = n_distinct(sample_id),
    species_sample_cells = n(),
    mean_absolute_error_pct_points =
      mean(abs(difference_pct_points)),
    root_mean_squared_error =
      sqrt(mean(difference_pct_points^2)),
    pearson_correlation = cor(
      `Expected: CFU-adjusted`,
      `Observed: custom Emu`,
      method = "pearson"
    ),
    spearman_correlation = cor(
      `Expected: CFU-adjusted`,
      `Observed: custom Emu`,
      method = "spearman"
    ),
    mixtures_passing_proposed_criteria =
      sum(
        known_mixture_validation$validation_status ==
          "Pass proposed semi-quantitative criteria"
      ),
    mixtures_reviewed =
      sum(
        known_mixture_validation$validation_status ==
          "Review"
      )
  )

kable(
  known_mixture_overall_stats,
  digits = 3,
  caption = "Overall known-mixture validation statistics"
)
```

## Draw the known-mixture validation figure

The dashed lines show the proposed exploratory criteria: MAE no greater than
10 percentage points and Bray-Curtis similarity of at least 0.80. These are
study-specific criteria, not universal thresholds.

```{r plot-known-mixture-validation, fig.width=11, fig.height=5.5}
known_mixture_plot_data <- known_mixture_validation %>%
  select(
    display_name,
    display_order,
    validation_status,
    mae_all_five_pct_points,
    bray_curtis_similarity
  ) %>%
  pivot_longer(
    cols = c(
      mae_all_five_pct_points,
      bray_curtis_similarity
    ),
    names_to = "metric",
    values_to = "metric_value"
  ) %>%
  mutate(
    display_name = factor(
      display_name,
      levels = known_mixture_validation$display_name
    ),
    metric = recode(
      metric,
      mae_all_five_pct_points =
        "MAE across all five species (percentage points)",
      bray_curtis_similarity =
        "Bray-Curtis similarity"
    )
  )

known_mixture_thresholds <- tibble(
  metric = c(
    "MAE across all five species (percentage points)",
    "Bray-Curtis similarity"
  ),
  threshold = c(10, 0.80)
)

validation_status_colors <- c(
  "Pass proposed semi-quantitative criteria" = "#009E73",
  "Review" = "#D55E00"
)

known_mixture_validation_plot <- ggplot(
  known_mixture_plot_data,
  aes(
    x = display_name,
    y = metric_value,
    fill = validation_status
  )
) +
  geom_col(width = 0.72) +
  geom_hline(
    data = known_mixture_thresholds,
    aes(yintercept = threshold),
    inherit.aes = FALSE,
    linetype = 2,
    color = "#404040"
  ) +
  geom_text(
    aes(
      label = if_else(
        metric == "Bray-Curtis similarity",
        number(metric_value, accuracy = 0.001),
        number(metric_value, accuracy = 0.1)
      )
    ),
    vjust = -0.35,
    size = 3.3
  ) +
  facet_wrap(
    vars(metric),
    scales = "free_y",
    labeller = label_wrap_gen(width = 28)
  ) +
  scale_fill_manual(values = validation_status_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    x = NULL,
    y = "Metric value",
    fill = NULL,
    title = "Known-mixture recovery",
    subtitle = "Dashed lines show proposed study-specific criteria"
  ) +
  theme_16s +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    strip.text = element_text(face = "bold")
  )

known_mixture_validation_plot
```

```{r save-known-mixture-validation}
write_tsv(
  known_mixture_validation,
  file.path(table_dir, "known_mixture_validation_stats.tsv")
)
write_tsv(
  known_mixture_overall_stats,
  file.path(table_dir, "known_mixture_overall_stats.tsv")
)

ggsave(
  file.path(figure_dir, "known_mixture_validation_stats.png"),
  known_mixture_validation_plot,
  width = 11,
  height = 5.5,
  dpi = 300
)
ggsave(
  file.path(figure_dir, "known_mixture_validation_stats.pdf"),
  known_mixture_validation_plot,
  width = 11,
  height = 5.5
)
```

# 9. Percentage of reads assigned by custom versus broad Emu

Here, custom Emu assignment means the percentage of QC-passing reads assigned
to one of the five custom references. Broad Emu assignment means the
percentage receiving any taxonomic assignment in the published database. The
table also retains the percentage that broad Emu assigned to the five exact
expected species.

## Calculate assignment percentages

```{r prepare-emu-assignment-rate}
custom_assignment_rate <- custom_raw %>%
  filter(sample_id %in% selected_sample_ids) %>%
  group_by(sample_id) %>%
  summarise(
    custom_five_reference_assigned_pct =
      sum(emu_abundance_pct_all_usable_reads),
    .groups = "drop"
  )

broad_assignment_rate <- broad_accounting_raw %>%
  filter(sample_id %in% selected_sample_ids) %>%
  transmute(
    sample_id,
    broad_any_taxon_assigned_pct =
      100 * emu_classified_estimated_reads / usable_reads,
    broad_five_exact_species_assigned_pct =
      100 * expected_species_estimated_reads / usable_reads
  )

emu_assignment_rate_table <- custom_assignment_rate %>%
  inner_join(broad_assignment_rate, by = "sample_id") %>%
  left_join(sample_key, by = "sample_id") %>%
  mutate(
    display_name = factor(display_name, levels = selected_sample_names),
    broad_minus_custom_pct_points =
      broad_any_taxon_assigned_pct -
      custom_five_reference_assigned_pct
  ) %>%
  arrange(display_order)

kable(
  emu_assignment_rate_table %>%
    select(
      display_name,
      custom_five_reference_assigned_pct,
      broad_any_taxon_assigned_pct,
      broad_five_exact_species_assigned_pct,
      broad_minus_custom_pct_points
    ),
  digits = 3,
  caption = "Custom and broad Emu assignment percentages"
)
```

## Summarize assignment rates

```{r summarize-emu-assignment-rate}
emu_assignment_rate_long <- emu_assignment_rate_table %>%
  select(
    sample_id,
    display_name,
    display_order,
    custom_five_reference_assigned_pct,
    broad_any_taxon_assigned_pct
  ) %>%
  pivot_longer(
    cols = c(
      custom_five_reference_assigned_pct,
      broad_any_taxon_assigned_pct
    ),
    names_to = "method",
    values_to = "assigned_pct"
  ) %>%
  mutate(
    method = recode(
      method,
      custom_five_reference_assigned_pct =
        "Custom Emu: five references",
      broad_any_taxon_assigned_pct =
        "Broad Emu: any classified taxon"
    ),
    method = factor(method, levels = names(assignment_method_colors))
  )

emu_assignment_method_stats <- emu_assignment_rate_long %>%
  group_by(method) %>%
  summarise(
    samples = n(),
    mean_assigned_pct = mean(assigned_pct),
    median_assigned_pct = median(assigned_pct),
    sd_assigned_pct = sd(assigned_pct),
    minimum_assigned_pct = min(assigned_pct),
    maximum_assigned_pct = max(assigned_pct),
    .groups = "drop"
  )

emu_assignment_paired_stats <- emu_assignment_rate_table %>%
  summarise(
    mean_broad_minus_custom_pct_points =
      mean(broad_minus_custom_pct_points),
    median_broad_minus_custom_pct_points =
      median(broad_minus_custom_pct_points),
    minimum_broad_minus_custom_pct_points =
      min(broad_minus_custom_pct_points),
    maximum_broad_minus_custom_pct_points =
      max(broad_minus_custom_pct_points)
  )

kable(
  emu_assignment_method_stats,
  digits = 3,
  caption = "Distribution of Emu assignment rates across selected samples"
)

kable(
  emu_assignment_paired_stats,
  digits = 3,
  caption = "Paired broad-minus-custom assignment difference"
)
```

## Draw the custom-versus-broad assignment graph

The point scale is zoomed because both methods assign most QC-passing reads.
The graph uses points rather than bars so the nonzero baseline does not imply
that the assigned fractions start at zero.

```{r plot-emu-assignment-rate, fig.width=12, fig.height=5.5}
assignment_y_min <- max(
  0,
  2 * floor(
    (min(emu_assignment_rate_long$assigned_pct) - 1) / 2
  )
)

emu_assignment_rate_plot <- ggplot(
  emu_assignment_rate_table,
  aes(x = display_name)
) +
  geom_segment(
    aes(
      xend = display_name,
      y = custom_five_reference_assigned_pct,
      yend = broad_any_taxon_assigned_pct
    ),
    color = "#B0B0B0",
    linewidth = 0.6
  ) +
  geom_point(
    aes(
      y = custom_five_reference_assigned_pct,
      color = "Custom Emu: five references"
    ),
    size = 2.8
  ) +
  geom_point(
    aes(
      y = broad_any_taxon_assigned_pct,
      color = "Broad Emu: any classified taxon"
    ),
    size = 2.8
  ) +
  scale_color_manual(values = assignment_method_colors, drop = FALSE) +
  scale_y_continuous(
    breaks = pretty_breaks(n = 6),
    labels = label_number(suffix = "%")
  ) +
  coord_cartesian(ylim = c(assignment_y_min, 100)) +
  labs(
    x = NULL,
    y = "QC-passing reads assigned",
    color = NULL,
    title = "Read assignment: custom versus broad Emu",
    subtitle = "Gray lines connect the two database results for each sample"
  ) +
  theme_16s

emu_assignment_rate_plot
```

```{r save-emu-assignment-rate}
write_tsv(
  emu_assignment_rate_table %>%
    mutate(display_name = as.character(display_name)),
  file.path(table_dir, "custom_vs_broad_emu_assignment_rates.tsv")
)
write_tsv(
  emu_assignment_method_stats,
  file.path(table_dir, "custom_vs_broad_emu_assignment_summary_stats.tsv")
)

ggsave(
  file.path(figure_dir, "custom_vs_broad_emu_assignment_rates.png"),
  emu_assignment_rate_plot,
  width = 12,
  height = 5.5,
  dpi = 300
)
ggsave(
  file.path(figure_dir, "custom_vs_broad_emu_assignment_rates.pdf"),
  emu_assignment_rate_plot,
  width = 12,
  height = 5.5
)
```

# 10. Minimap2 read-level cross-check

Minimap2 classifies each filtered read as `confident`, `ambiguous`, or
`unassigned`. A confident read passes the configured alignment-length,
coverage, identity, mapping-quality, and best-versus-second-score thresholds.
Ambiguous reads have two closely scoring reference alignments and are not
forced onto one species.

## Prepare minimap2 read-status data

```{r prepare-minimap-status}
minimap_status_table <- minimap_status_raw %>%
  filter(sample_id %in% selected_sample_ids) %>%
  left_join(sample_key, by = "sample_id") %>%
  mutate(display_name = factor(display_name, levels = selected_sample_names)) %>%
  arrange(display_order)

minimap_status_long <- minimap_status_table %>%
  select(
    sample_id,
    display_name,
    display_order,
    confident_pct,
    ambiguous_pct,
    unassigned_pct
  ) %>%
  pivot_longer(
    cols = ends_with("_pct"),
    names_to = "assignment_status",
    values_to = "read_pct"
  ) %>%
  mutate(
    assignment_status = recode(
      assignment_status,
      confident_pct = "Confident",
      ambiguous_pct = "Ambiguous",
      unassigned_pct = "Unassigned"
    ),
    assignment_status = factor(
      assignment_status,
      levels = names(minimap_status_colors)
    )
  )

kable(
  minimap_status_table %>%
    select(
      display_name,
      total_reads,
      confident_reads,
      ambiguous_reads,
      unassigned_reads,
      confident_pct,
      ambiguous_pct,
      unassigned_pct
    ),
  digits = 3,
  caption = "Minimap2 read-status results"
)
```

## Display overall and across-sample minimap2 statistics

```{r summarize-minimap-status}
minimap_weighted_stats <- minimap_status_table %>%
  summarise(
    selected_samples = n(),
    total_reads = sum(total_reads),
    confident_reads = sum(confident_reads),
    ambiguous_reads = sum(ambiguous_reads),
    unassigned_reads = sum(unassigned_reads),
    confident_pct = 100 * confident_reads / total_reads,
    ambiguous_pct = 100 * ambiguous_reads / total_reads,
    unassigned_pct = 100 * unassigned_reads / total_reads
  )

minimap_status_distribution_stats <- minimap_status_long %>%
  group_by(assignment_status) %>%
  summarise(
    mean_pct = mean(read_pct),
    median_pct = median(read_pct),
    sd_pct = sd(read_pct),
    minimum_pct = min(read_pct),
    maximum_pct = max(read_pct),
    .groups = "drop"
  )

kable(
  minimap_weighted_stats,
  digits = 3,
  caption = "Read-weighted minimap2 statistics across selected samples"
)

kable(
  minimap_status_distribution_stats,
  digits = 3,
  caption = "Distribution of minimap2 percentages across samples"
)
```

## Draw per-sample minimap2 status plots

Each panel has its own y-axis range so the small ambiguous fraction remains
visible.

```{r plot-minimap-status, fig.width=12, fig.height=8}
minimap_status_plot <- ggplot(
  minimap_status_long,
  aes(x = display_name, y = read_pct, fill = assignment_status)
) +
  geom_col(width = 0.78) +
  facet_wrap(
    vars(assignment_status),
    ncol = 1,
    scales = "free_y"
  ) +
  scale_fill_manual(values = minimap_status_colors, drop = FALSE) +
  scale_y_continuous(
    labels = label_number(suffix = "%"),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    x = NULL,
    y = "Filtered reads",
    fill = NULL,
    title = "Minimap2 read-level assignment status",
    subtitle = "Independent alignment against the five custom references"
  ) +
  theme_16s +
  theme(legend.position = "none")

minimap_status_plot
```

```{r save-minimap-status}
write_tsv(
  minimap_status_table %>%
    mutate(display_name = as.character(display_name)),
  file.path(table_dir, "minimap2_status_selected_samples.tsv")
)
write_tsv(
  minimap_weighted_stats,
  file.path(table_dir, "minimap2_weighted_status_stats.tsv")
)
write_tsv(
  minimap_status_distribution_stats,
  file.path(table_dir, "minimap2_status_distribution_stats.tsv")
)

ggsave(
  file.path(figure_dir, "minimap2_read_status_selected_samples.png"),
  minimap_status_plot,
  width = 12,
  height = 8,
  dpi = 300
)
ggsave(
  file.path(figure_dir, "minimap2_read_status_selected_samples.pdf"),
  minimap_status_plot,
  width = 12,
  height = 8
)
```

## Compare custom Emu with confident minimap2 assignments

Both percentages in this comparison are renormalized among the five expected
species. Minimap2 uses only reads classified as confident.

```{r prepare-custom-minimap-comparison}
minimap_composition <- minimap_confident_raw %>%
  filter(sample_id %in% selected_sample_ids) %>%
  transmute(
    sample_id,
    species = reference %>%
      str_remove("_ref$") %>%
      str_replace_all("_", " "),
    minimap_pct = percentage_among_confident_reads
  ) %>%
  complete(
    sample_id = selected_sample_ids,
    species = species_order,
    fill = list(minimap_pct = 0)
  )

custom_minimap_comparison <- custom_plot_data %>%
  transmute(
    sample_id,
    display_name = as.character(display_name),
    display_order,
    species = as.character(species),
    custom_emu_pct = abundance_pct
  ) %>%
  left_join(
    minimap_composition,
    by = c("sample_id", "species")
  ) %>%
  mutate(
    difference_pct_points = minimap_pct - custom_emu_pct,
    absolute_difference_pct_points = abs(difference_pct_points)
  )

agreement_stats <- function(data, scope_label) {
  data %>%
    summarise(
      comparison_cells = n(),
      mean_absolute_error_pct_points =
        mean(absolute_difference_pct_points),
      root_mean_squared_error =
        sqrt(mean(difference_pct_points^2)),
      pearson_correlation =
        cor(custom_emu_pct, minimap_pct, method = "pearson"),
      spearman_correlation =
        cor(custom_emu_pct, minimap_pct, method = "spearman"),
      maximum_absolute_difference_pct_points =
        max(absolute_difference_pct_points)
    ) %>%
    mutate(comparison_scope = scope_label, .before = 1)
}

custom_minimap_agreement_stats <- bind_rows(
  agreement_stats(
    custom_minimap_comparison,
    "All five species in all selected samples"
  ),
  agreement_stats(
    custom_minimap_comparison %>%
      filter(custom_emu_pct > 0 | minimap_pct > 0),
    "Species detected by either method"
  )
)

custom_minimap_sample_stats <- custom_minimap_comparison %>%
  group_by(sample_id, display_name, display_order) %>%
  summarise(
    mean_absolute_error_pct_points =
      mean(absolute_difference_pct_points),
    maximum_absolute_difference_pct_points =
      max(absolute_difference_pct_points),
    .groups = "drop"
  ) %>%
  arrange(display_order)

kable(
  custom_minimap_agreement_stats,
  digits = 4,
  caption = "Custom Emu versus minimap2 agreement statistics"
)

kable(
  custom_minimap_sample_stats %>% select(-display_order),
  digits = 3,
  caption = "Per-sample custom Emu versus minimap2 differences"
)
```

## Draw custom Emu versus minimap2 agreement

```{r plot-custom-minimap-comparison, fig.width=11, fig.height=7}
custom_minimap_plot <- ggplot(
  custom_minimap_comparison,
  aes(
    x = custom_emu_pct,
    y = minimap_pct,
    color = species
  )
) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = 2,
    color = "#606060"
  ) +
  geom_point(size = 2.3, alpha = 0.8) +
  facet_wrap(
    vars(species),
    labeller = label_wrap_gen(width = 24)
  ) +
  scale_color_manual(values = species_colors[species_order], drop = FALSE) +
  scale_x_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, by = 20),
    labels = label_number(suffix = "%")
  ) +
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, by = 20),
    labels = label_number(suffix = "%")
  ) +
  coord_equal() +
  labs(
    x = "Custom Emu composition",
    y = "Confident minimap2 composition",
    color = NULL,
    title = "Custom Emu versus minimap2",
    subtitle = "Dashed line indicates exact agreement"
  ) +
  theme_16s +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )

custom_minimap_plot
```

```{r save-custom-minimap-comparison}
write_tsv(
  custom_minimap_comparison,
  file.path(table_dir, "custom_emu_vs_minimap2_selected_samples.tsv")
)
write_tsv(
  custom_minimap_agreement_stats,
  file.path(table_dir, "custom_emu_vs_minimap2_agreement_stats.tsv")
)
write_tsv(
  custom_minimap_sample_stats,
  file.path(table_dir, "custom_emu_vs_minimap2_per_sample_stats.tsv")
)

ggsave(
  file.path(figure_dir, "custom_emu_vs_minimap2_selected_samples.png"),
  custom_minimap_plot,
  width = 11,
  height = 7,
  dpi = 300
)
ggsave(
  file.path(figure_dir, "custom_emu_vs_minimap2_selected_samples.pdf"),
  custom_minimap_plot,
  width = 11,
  height = 7
)
```

# 11. Optional: inspect specific broad Emu unexpected taxa

This table identifies the taxa behind the assigned portion of
`Unknown / other`. Technical categories such as unmapped and
alignment-filtered reads are not species and are shown separately in Section 6.

```{r inspect-unexpected-taxa}
broad_unexpected_taxa <- broad_species_raw %>%
  filter(
    sample_id %in% selected_sample_ids,
    !species %in% species_order
  ) %>%
  left_join(sample_key, by = "sample_id") %>%
  select(
    display_order,
    display_name,
    tax_id,
    species,
    genus,
    abundance_pct_all_usable_reads,
    estimated_counts
  ) %>%
  arrange(display_order, desc(abundance_pct_all_usable_reads))

kable(
  broad_unexpected_taxa %>% select(-display_order),
  digits = 3,
  caption = "Unexpected species assigned by broad Emu"
)
```

# 12. P1-normalized organism trajectories

This section follows the five expected organisms through P1, P3, P5, P7, and
P9 using custom Emu abundance among **all QC-passing reads**. Two baseline
transformations are shown:

For this longitudinal analysis, P1 is sample 16 collected at 48 hours because
that culture was used for passaging. Sample 1 was collected at 24 hours and is
not used as the passage baseline. Its results remain in the project and in
other applicable cross-sectional plots.

Sample 16 had an atypical raw read-length distribution (raw median 1,060 bp).
After the standard 1,200--1,800 bp and Q10 filters, 1,977 reads remained with
a median length of 1,496 bp. Its use here is therefore documented explicitly
rather than treated as an interchangeable replacement based only on read
count.

1. **Change from P1 in percentage points** equals the abundance at each point
   minus its P1 abundance. P1 is therefore zero for every organism. This is the
   primary view because it remains defined when an organism was absent at P1.
2. **P1-normalized index** equals `100 * abundance / P1 abundance`. P1 equals
   100, 200 means twice the P1 relative abundance, and 50 means half the P1
   relative abundance. It is undefined when P1 abundance is zero.

These remain compositional measurements. They show how each organism's share
of the sequenced community changed; they do not demonstrate absolute growth or
decline without an independent total-load measurement such as total CFU,
16S-qPCR, or spike-in-normalized abundance.

## Define the P1-P9 series

```{r define-p1-p9-series}
trajectory_key_file <- path(
  "metadata",
  "longitudinal_trajectory_samples.tsv"
)
if (!file.exists(trajectory_key_file)) {
  stop("Missing longitudinal trajectory metadata: ", trajectory_key_file)
}

trajectory_key <- read_tsv(
  trajectory_key_file,
  show_col_types = FALSE
) %>%
  arrange(trajectory_order)

if (anyDuplicated(trajectory_key$sample_id)) {
  stop("Duplicate sample IDs in longitudinal trajectory metadata")
}
if (anyDuplicated(trajectory_key$trajectory_point)) {
  stop("Duplicate trajectory points in longitudinal trajectory metadata")
}
if (!identical(trajectory_key$trajectory_point, c("P1", "P3", "P5", "P7", "P9"))) {
  stop("Longitudinal trajectory points must be ordered P1, P3, P5, P7, P9")
}

trajectory_custom_raw <- custom_raw %>%
  select(
    sample_id,
    species,
    emu_abundance_pct_renormalized_among_expected,
    emu_abundance_pct_all_usable_reads,
    estimated_counts
  )

missing_trajectory_samples <- setdiff(
  trajectory_key$sample_id,
  unique(trajectory_custom_raw$sample_id)
)

if (length(missing_trajectory_samples) > 0) {
  stop(
    "P1-P9 sample(s) missing from custom Emu results: ",
    paste(missing_trajectory_samples, collapse = ", ")
  )
}

kable(
  trajectory_key,
  caption = "Samples used for the P1-P9 passage trajectory"
)
```

## Passage-only custom Emu composition

This graph contains only the longitudinal passage series. Percentages are
renormalized among reads assigned to the five expected species, so every bar
sums to 100%. Use the P1-centered graphs below to inspect organism-specific
changes while retaining the all-usable-read denominator.

```{r prepare-passage-only-stacked}
passage_custom_plot_data <- trajectory_custom_raw %>%
  filter(sample_id %in% trajectory_key$sample_id) %>%
  select(
    sample_id,
    species,
    abundance_pct =
      emu_abundance_pct_renormalized_among_expected
  ) %>%
  complete(
    sample_id = trajectory_key$sample_id,
    species = species_order,
    fill = list(abundance_pct = 0)
  ) %>%
  left_join(trajectory_key, by = "sample_id") %>%
  mutate(
    trajectory_point = factor(
      trajectory_point,
      levels = trajectory_key$trajectory_point
    ),
    species = factor(species, levels = species_order)
  ) %>%
  arrange(trajectory_order, species)

kable(
  passage_custom_plot_data %>%
    mutate(
      trajectory_point = as.character(trajectory_point),
      species = as.character(species)
    ) %>%
    select(trajectory_point, species, abundance_pct),
  digits = 3,
  caption = "Custom Emu composition for the passage-only series"
)
```

```{r plot-passage-only-stacked, fig.width=9, fig.height=5.5}
passage_custom_stacked_plot <- ggplot(
  passage_custom_plot_data,
  aes(
    x = trajectory_point,
    y = abundance_pct,
    fill = species
  )
) +
  geom_col(width = 0.78, color = "white", linewidth = 0.2) +
  scale_fill_manual(
    values = species_colors[species_order],
    drop = FALSE
  ) +
  guides(
    fill = guide_legend(nrow = 2, byrow = TRUE)
  ) +
  scale_y_continuous(
    breaks = seq(0, 100, by = 20),
    labels = label_number(suffix = "%"),
    expand = expansion(mult = c(0, 0))
  ) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(
    x = "Passage",
    y = "Abundance among five expected species",
    fill = NULL,
    title = "Community composition across passages",
    subtitle = "Custom five-reference Emu; P1 is the 48h passage baseline"
  ) +
  theme_16s +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.box = "vertical"
  )

passage_custom_stacked_plot
```

```{r save-passage-only-stacked}
write_tsv(
  passage_custom_plot_data %>%
    mutate(
      trajectory_point = as.character(trajectory_point),
      species = as.character(species)
    ),
  file.path(table_dir, "passage_only_custom_emu_stacked_plot_data.tsv")
)

ggsave(
  file.path(figure_dir, "passage_only_custom_emu_stacked.png"),
  passage_custom_stacked_plot,
  width = 9,
  height = 5.5,
  dpi = 300
)
ggsave(
  file.path(figure_dir, "passage_only_custom_emu_stacked.pdf"),
  passage_custom_stacked_plot,
  width = 9,
  height = 5.5
)
```

## Calculate P1-centered changes and indices

```{r prepare-p1-normalized-trajectories}
trajectory_abundance <- trajectory_custom_raw %>%
  filter(sample_id %in% trajectory_key$sample_id) %>%
  select(
    sample_id,
    species,
    abundance_pct_all_usable_reads =
      emu_abundance_pct_all_usable_reads,
    estimated_counts
  ) %>%
  complete(
    sample_id = trajectory_key$sample_id,
    species = species_order,
    fill = list(
      abundance_pct_all_usable_reads = 0,
      estimated_counts = 0
    )
  ) %>%
  left_join(trajectory_key, by = "sample_id")

p1_baseline <- trajectory_abundance %>%
  filter(trajectory_point == "P1") %>%
  transmute(
    species,
    p1_abundance_pct = abundance_pct_all_usable_reads,
    p1_estimated_counts = estimated_counts,
    p1_baseline_support = case_when(
      p1_estimated_counts >= 10 ~
        "Supported: at least 10 estimated P1 reads",
      p1_estimated_counts > 0 ~
        "Low support: fewer than 10 estimated P1 reads",
      TRUE ~
        "Not detected at P1: fold index undefined"
    )
  )

trajectory_normalized <- trajectory_abundance %>%
  left_join(p1_baseline, by = "species") %>%
  mutate(
    change_from_p1_pct_points =
      abundance_pct_all_usable_reads - p1_abundance_pct,
    p1_normalized_index = if_else(
      p1_abundance_pct > 0,
      100 * abundance_pct_all_usable_reads / p1_abundance_pct,
      NA_real_
    ),
    species = factor(species, levels = species_order),
    trajectory_point = factor(
      trajectory_point,
      levels = trajectory_key$trajectory_point
    )
  ) %>%
  arrange(species, trajectory_order)

p1_usable_reads <- sample16_custom_total_reads

kable(
  p1_baseline %>%
    mutate(p1_total_usable_reads = p1_usable_reads) %>%
    select(
      species,
      p1_total_usable_reads,
      p1_estimated_counts,
      p1_abundance_pct,
      p1_baseline_support
    ),
  digits = 3,
  caption = "P1 baseline support for fold-change interpretation"
)
```

## Inspect the normalized trajectory table

```{r display-p1-normalized-trajectories}
kable(
  trajectory_normalized %>%
    mutate(species = as.character(species)) %>%
    select(
      trajectory_point,
      species,
      abundance_pct_all_usable_reads,
      change_from_p1_pct_points,
      p1_normalized_index,
      p1_estimated_counts,
      p1_baseline_support
    ),
  digits = 3,
  caption = "P1-normalized organism trajectories"
)
```

## Draw change from P1 for every organism

The panels use different y-axis ranges so changes in low-abundance organisms
remain visible. Compare numerical values, rather than bar or panel height,
when comparing the magnitude of change between organisms.

```{r plot-change-from-p1, fig.width=11, fig.height=7}
change_from_p1_plot <- ggplot(
  trajectory_normalized,
  aes(
    x = trajectory_order,
    y = change_from_p1_pct_points,
    color = species,
    group = species
  )
) +
  geom_hline(
    yintercept = 0,
    linetype = 2,
    color = "#606060"
  ) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.8) +
  facet_wrap(
    vars(species),
    scales = "free_y",
    labeller = label_wrap_gen(width = 24)
  ) +
  scale_color_manual(values = species_colors[species_order], drop = FALSE) +
  scale_x_continuous(
    breaks = trajectory_key$trajectory_order,
    labels = trajectory_key$trajectory_point
  ) +
  scale_y_continuous(
    labels = label_number(suffix = " pp"),
    expand = expansion(mult = c(0.12, 0.12))
  ) +
  labs(
    x = "Sampling point",
    y = "Change from P1",
    color = NULL,
    title = "Organism trajectories centered on P1",
    subtitle = "Custom Emu abundance among all QC-passing reads"
  ) +
  theme_16s +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )

change_from_p1_plot
```

## Draw the P1-normalized index

An organism with zero estimated P1 reads has an undefined fold index and is
annotated in its panel. Interpret large indices cautiously whenever the P1
baseline has low read support.

```{r plot-p1-normalized-index, fig.width=11, fig.height=9}
p1_index_annotation <- p1_baseline %>%
  filter(p1_abundance_pct == 0) %>%
  transmute(
    species = factor(species, levels = species_order),
    trajectory_order = 5,
    p1_normalized_index = 100,
    label = "P1 = 0 reads\nfold index undefined"
  )

p1_normalized_index_plot <- ggplot(
  trajectory_normalized,
  aes(
    x = trajectory_order,
    y = p1_normalized_index,
    color = species,
    group = species
  )
) +
  geom_hline(
    yintercept = 100,
    linetype = 2,
    color = "#606060"
  ) +
  geom_line(linewidth = 0.8, na.rm = TRUE) +
  geom_point(size = 2.8, na.rm = TRUE) +
  geom_text(
    data = p1_index_annotation,
    aes(
      x = trajectory_order,
      y = p1_normalized_index,
      label = label
    ),
    inherit.aes = FALSE,
    size = 3.4
  ) +
  facet_wrap(
    vars(species),
    ncol = 2,
    scales = "free_y",
    drop = FALSE,
    labeller = label_wrap_gen(width = 24)
  ) +
  scale_color_manual(values = species_colors[species_order], drop = FALSE) +
  scale_x_continuous(
    breaks = trajectory_key$trajectory_order,
    labels = trajectory_key$trajectory_point
  ) +
  scale_y_continuous(
    labels = label_number(),
    expand = expansion(mult = c(0.12, 0.15))
  ) +
  labs(
    x = "Sampling point",
    y = "Relative-composition index (not reads)",
    color = NULL,
    title = "Relative composition indexed to P1 = 100",
    subtitle = "P1 = 100 is a ratio baseline; it is not a read count"
  ) +
  theme_16s +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )

p1_normalized_index_plot
```

```{r save-p1-normalized-trajectories}
write_tsv(
  trajectory_normalized %>%
    mutate(
      species = as.character(species),
      trajectory_point = as.character(trajectory_point)
    ),
  file.path(table_dir, "p1_normalized_organism_trajectories.tsv")
)
write_tsv(
  p1_baseline,
  file.path(table_dir, "p1_baseline_support.tsv")
)

ggsave(
  file.path(figure_dir, "organism_change_from_p1.png"),
  change_from_p1_plot,
  width = 11,
  height = 7,
  dpi = 300
)
ggsave(
  file.path(figure_dir, "organism_change_from_p1.pdf"),
  change_from_p1_plot,
  width = 11,
  height = 7
)
ggsave(
  file.path(figure_dir, "organism_p1_normalized_index.png"),
  p1_normalized_index_plot,
  width = 11,
  height = 9,
  dpi = 300
)
ggsave(
  file.path(figure_dir, "organism_p1_normalized_index.pdf"),
  p1_normalized_index_plot,
  width = 11,
  height = 9
)
```

# 13. P5 to P6 frozen-stock branch

This section tracks the three organisms detected across P5, the P6 frozen
stock, and the two cultures started from that stock. The two branch lines
share P5 and the frozen-stock measurement, then separate at `P6_p1_24` and
`P6_p1_48`.

The plotted value is change from P5 in **percentage points**, calculated from
custom Emu abundance among all QC-passing reads. This is a relative-composition
measurement, not absolute growth. The accompanying table retains the observed
percentages and estimated read counts.

## Define the branch and calculate change from P5

```{r prepare-p5-p6-branch}
p6_branch_species <- c(
  "Bacteroides thetaiotaomicron",
  "Fusobacterium nucleatum",
  "Veillonella parvula"
)

p6_branch_key <- tribble(
  ~sample_id,              ~branch_stage,       ~stage_order, ~branch_endpoint,
  "2PJGZM_18_sample_18",   "P5",                           1, "P6_p1_24",
  "2PJGZM_10_sample_10",   "P6 frozen stock",              2, "P6_p1_24",
  "2PJGZM_8_sample_8",     "P6_p1_24",                     3, "P6_p1_24",
  "2PJGZM_18_sample_18",   "P5",                           1, "P6_p1_48",
  "2PJGZM_10_sample_10",   "P6 frozen stock",              2, "P6_p1_48",
  "2PJGZM_3_sample_3",     "P6_p1_48",                     3, "P6_p1_48"
)

missing_p6_branch_samples <- setdiff(
  unique(p6_branch_key$sample_id),
  unique(custom_raw$sample_id)
)
if (length(missing_p6_branch_samples) > 0) {
  stop(
    "P5-P6 branch sample(s) missing from custom Emu results: ",
    paste(missing_p6_branch_samples, collapse = ", ")
  )
}

p5_branch_baseline <- custom_raw %>%
  filter(
    sample_id == "2PJGZM_18_sample_18",
    species %in% p6_branch_species
  ) %>%
  transmute(
    species,
    p5_abundance_pct = emu_abundance_pct_all_usable_reads,
    p5_estimated_counts = estimated_counts
  )

if (
  nrow(p5_branch_baseline) != length(p6_branch_species) ||
    any(p5_branch_baseline$p5_estimated_counts <= 0)
) {
  stop("P5 does not contain supported baselines for all three branch species")
}

p6_branch_tracking <- custom_raw %>%
  filter(
    sample_id %in% unique(p6_branch_key$sample_id),
    species %in% p6_branch_species
  ) %>%
  select(
    sample_id,
    species,
    abundance_pct_all_usable_reads =
      emu_abundance_pct_all_usable_reads,
    estimated_counts
  ) %>%
  inner_join(p6_branch_key, by = "sample_id") %>%
  left_join(p5_branch_baseline, by = "species") %>%
  mutate(
    change_from_p5_pct_points =
      abundance_pct_all_usable_reads - p5_abundance_pct,
    species = factor(species, levels = p6_branch_species),
    branch_endpoint = factor(
      branch_endpoint,
      levels = c("P6_p1_24", "P6_p1_48")
    )
  ) %>%
  arrange(species, branch_endpoint, stage_order)

p6_branch_display_table <- p6_branch_tracking %>%
  distinct(
    sample_id,
    branch_stage,
    stage_order,
    species,
    abundance_pct_all_usable_reads,
    estimated_counts,
    p5_abundance_pct,
    change_from_p5_pct_points
  ) %>%
  arrange(stage_order, species)

kable(
  p6_branch_display_table %>%
    select(
      branch_stage,
      species,
      abundance_pct_all_usable_reads,
      estimated_counts,
      change_from_p5_pct_points
    ),
  digits = 3,
  caption = "Three-species composition along the P5-P6 branch"
)
```

## Draw the shared path and parallel branches

```{r plot-p5-p6-branch, fig.width=10, fig.height=6}
p6_branch_plot <- ggplot(
  p6_branch_tracking,
  aes(
    x = stage_order,
    y = change_from_p5_pct_points,
    color = species,
    linetype = branch_endpoint,
    shape = branch_endpoint,
    group = interaction(species, branch_endpoint)
  )
) +
  geom_hline(
    yintercept = 0,
    linetype = 3,
    color = "#606060"
  ) +
  geom_line(linewidth = 1) +
  geom_point(
    data = p6_branch_tracking %>%
      filter(stage_order < 3) %>%
      distinct(species, stage_order, .keep_all = TRUE),
    aes(
      x = stage_order,
      y = change_from_p5_pct_points,
      color = species
    ),
    inherit.aes = FALSE,
    shape = 16,
    size = 3.2
  ) +
  geom_point(
    data = p6_branch_tracking %>%
      filter(stage_order == 3),
    size = 3.2,
    stroke = 0.9
  ) +
  scale_color_manual(
    values = species_colors[p6_branch_species],
    drop = FALSE
  ) +
  scale_linetype_manual(
    values = c(
      "P6_p1_24" = "solid",
      "P6_p1_48" = "22"
    ),
    labels = c(
      "P6_p1_24" = "Branch to P6_p1_24",
      "P6_p1_48" = "Branch to P6_p1_48"
    )
  ) +
  scale_shape_manual(
    values = c(
      "P6_p1_24" = 16,
      "P6_p1_48" = 17
    ),
    labels = c(
      "P6_p1_24" = "Branch to P6_p1_24",
      "P6_p1_48" = "Branch to P6_p1_48"
    )
  ) +
  scale_x_continuous(
    breaks = c(1, 2, 3),
    labels = c(
      "P5",
      "P6 frozen stock",
      "P6_p1 endpoints"
    ),
    expand = expansion(mult = c(0.06, 0.08))
  ) +
  scale_y_continuous(
    labels = label_number(suffix = " pp"),
    expand = expansion(mult = c(0.12, 0.12))
  ) +
  labs(
    x = NULL,
    y = "Change from P5",
    color = NULL,
    linetype = NULL,
    shape = NULL,
    title = "Three-species change from P5 through the P6 frozen stock",
    subtitle = paste(
      "Shared path branches to P6_p1_24 and P6_p1_48;",
      "custom Emu percentage among all QC-passing reads"
    )
  ) +
  theme_16s +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.box = "vertical"
  ) +
  guides(
    color = guide_legend(order = 1, nrow = 2, byrow = TRUE),
    linetype = guide_legend(order = 2),
    shape = guide_legend(order = 2)
  )

p6_branch_plot
```

```{r save-p5-p6-branch}
write_tsv(
  p6_branch_tracking %>%
    mutate(
      species = as.character(species),
      branch_endpoint = as.character(branch_endpoint)
    ),
  file.path(table_dir, "p5_p6_frozen_stock_branch_tracking.tsv")
)

ggsave(
  file.path(figure_dir, "p5_p6_frozen_stock_branch.png"),
  p6_branch_plot,
  width = 10,
  height = 6,
  dpi = 300
)
ggsave(
  file.path(figure_dir, "p5_p6_frozen_stock_branch.pdf"),
  p6_branch_plot,
  width = 10,
  height = 6
)
```

## Custom Emu composition of the three P6 samples

This stacked graph compares only the P6 frozen stock and the two cultures
started from it. Percentages are renormalized among reads assigned to the five
expected species, matching the main custom Emu stacked graph. Cutibacterium
and Prevotella remain in the fixed species order but have zero-height segments
in all three samples.

```{r prepare-p6-three-sample-stacked}
p6_stacked_key <- tribble(
  ~sample_id,              ~p6_display_name,    ~p6_display_order,
  "2PJGZM_10_sample_10",   "P6 frozen stock",                   1,
  "2PJGZM_8_sample_8",     "P6_p1_24",                          2,
  "2PJGZM_3_sample_3",     "P6_p1_48",                          3
)

missing_p6_stacked_samples <- setdiff(
  p6_stacked_key$sample_id,
  unique(custom_raw$sample_id)
)
if (length(missing_p6_stacked_samples) > 0) {
  stop(
    "P6 stacked-graph sample(s) missing from custom Emu results: ",
    paste(missing_p6_stacked_samples, collapse = ", ")
  )
}

p6_stacked_plot_data <- custom_raw %>%
  filter(sample_id %in% p6_stacked_key$sample_id) %>%
  select(
    sample_id,
    species,
    abundance_pct =
      emu_abundance_pct_renormalized_among_expected,
    estimated_counts
  ) %>%
  complete(
    sample_id = p6_stacked_key$sample_id,
    species = species_order,
    fill = list(
      abundance_pct = 0,
      estimated_counts = 0
    )
  ) %>%
  left_join(p6_stacked_key, by = "sample_id") %>%
  mutate(
    p6_display_name = factor(
      p6_display_name,
      levels = p6_stacked_key$p6_display_name
    ),
    species = factor(species, levels = species_order)
  ) %>%
  arrange(p6_display_order, species)

p6_stacked_sum_check <- p6_stacked_plot_data %>%
  group_by(sample_id, p6_display_name) %>%
  summarise(total_pct = sum(abundance_pct), .groups = "drop")

kable(
  p6_stacked_plot_data %>%
    mutate(
      p6_display_name = as.character(p6_display_name),
      species = as.character(species)
    ) %>%
    select(
      p6_display_name,
      species,
      abundance_pct,
      estimated_counts
    ),
  digits = 3,
  caption = "Custom Emu composition of the three P6 samples"
)

kable(
  p6_stacked_sum_check,
  digits = 3,
  caption = "P6 custom Emu percentage-sum check"
)
```

```{r plot-p6-three-sample-stacked, fig.width=8.5, fig.height=5.5}
p6_three_sample_stacked_plot <- ggplot(
  p6_stacked_plot_data,
  aes(
    x = p6_display_name,
    y = abundance_pct,
    fill = species
  )
) +
  geom_col(width = 0.72, color = "white", linewidth = 0.2) +
  scale_fill_manual(
    values = species_colors[species_order],
    drop = FALSE
  ) +
  guides(
    fill = guide_legend(nrow = 2, byrow = TRUE)
  ) +
  scale_y_continuous(
    breaks = seq(0, 100, by = 20),
    labels = label_number(suffix = "%"),
    expand = expansion(mult = c(0, 0))
  ) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(
    x = NULL,
    y = "Abundance among five expected species",
    fill = NULL,
    title = "Custom Emu composition of the three P6 samples",
    subtitle = "P6 frozen stock compared with P6_p1_24 and P6_p1_48"
  ) +
  theme_16s +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.box = "vertical"
  )

p6_three_sample_stacked_plot
```

```{r save-p6-three-sample-stacked}
write_tsv(
  p6_stacked_plot_data %>%
    mutate(
      p6_display_name = as.character(p6_display_name),
      species = as.character(species)
    ),
  file.path(table_dir, "p6_three_sample_custom_emu_stacked.tsv")
)

ggsave(
  file.path(figure_dir, "p6_three_sample_custom_emu_stacked.png"),
  p6_three_sample_stacked_plot,
  width = 8.5,
  height = 5.5,
  dpi = 300
)
ggsave(
  file.path(figure_dir, "p6_three_sample_custom_emu_stacked.pdf"),
  p6_three_sample_stacked_plot,
  width = 8.5,
  height = 5.5
)
```

# 14. R session information

```{r session-information}
sessionInfo()
```
